---
title: "Assignment 11"
author: "Rachel Montgomery"
output: pdf_document
editor_options: 
  markdown: 
    wrap: sentence
---

# Analyzing the Impact of Artist Morality on Art Evaluations

In this blog post, we will explore a dataset collected from a survey that aimed to understand how people evaluate artworks and the potential influence of artists' morality on those evaluations.
We will walk through various statistical tests and analyses to answer the research question: Does an artist's morality affect evaluations of their artwork? 
In other words: Can people separate the art from the artist?

## Data Overview

The dataset comprises responses from 7,216 participants collected via Amazon Mechanical Turk (MTurk).
After applying screening for manipulation checks, we have a final sample size of 250 participants.
The data was collected through Qualtrics and includes various variables:

-   Participant: Unique participant identifiers.

-   FamousPerson: Names of famous individuals.

-   Painting: Identifiers for paintings.

-   Morality: Ratings of the morality of famous people.

-   Attractiveness: Ratings of the attractiveness of paintings.
    Beauty: Ratings of the beauty of paintings.

-   Skill: Ratings of the skill demonstrated in paintings.

-   Political: Political orientations of participants.

-   AEQ: Scores related to aesthetic experiences.

-   Age, Education, Race, Gender, SexualOrientation, SocialPolitical, EconomicPolitical: Demographic information.

## Setting up Environment and Data

First, let's set up our environment and load in the necessary libraries.

```{r message=FALSE, warning=FALSE}
# set seed for reproducibility
set.seed(16)

# load libraries
library(tidyverse)
library(rstatix)
```

Now we can read in our data.

```{r}
# load in data
load("BIU_Analysis_Data.RData")

art_morality <- mlm_df ##renaming data object so it makes more sense in our analysis  
```

## Analysis 

### 1. Write a function to perform a *t*-test

We begin by creating a custom t-test function to compare means between two groups.
A t-test of means is a statistical test that is used to compare the means of two groups to see if they are significantly different from each other. 

```{r, echo = TRUE, eval = TRUE, comment = NA, message = FALSE, warning = FALSE}
# creating function for t-test
t_test <- function(x1, x2)
{
  # calculate the mean of x1 and x2
  m1 <- mean(x1, na.rm = TRUE)
  m2 <- mean(x2, na.rm = TRUE)
  
  # calculate the variance of x1 and x2
  s1 <- var(x1, na.rm = TRUE)
  s2 <- var(x2, na.rm = TRUE)
  
  # calculate the sample sizes of x1 and x2
  n1 <- length(x1)
  n2 <- length(x2)
  
  # calculate the standard error of the difference in means
  se <- sqrt(s1 / n1 + s2 / n2)
  
  # calculate the t-statistic
  t_stat <- (m1 - m2) / se
  
  # calculate the degrees of freedom
  df <- n1 + n2 - 2
  
  # calculate the p-value
  p_value <- 2 * pt(abs(t_stat), df, lower.tail = FALSE)

  # create a result object with t-statistic and p-value
  result <- list(
    t_statistic = t_stat,
    degrees_of_freedom = df,
    p_value = p_value
  )
  
  # return the result
  return(result)
}
```

### 2. Write a function to compute Cohen's *d*

Next, we develop a function to compute Cohen's d, which measures the effect size between two groups.

Cohen's D is a statistical measure used to quantify the size of the effect in a study, often used in the context of comparing two means.
It's particularly common in the fields of psychology and education but is applicable in many areas of research where comparing group means is important.

```{r, echo = TRUE, eval = TRUE, comment = NA, message = FALSE, warning = FALSE}
# create function for Cohen's d
cohens_d <- function(x1, x2)
{
  # calculate the mean of x1 and x2
  m1 <- mean(x1, na.rm = TRUE)
  m2 <- mean(x2, na.rm = TRUE)
  
  # calculate the variance of x1 and x2
  s1 <- var(x1, na.rm = TRUE)
  s2 <- var(x2, na.rm = TRUE)
  
  # calculate the sample sizes of x1 and x2
  n1 <- length(x1)
  n2 <- length(x2)
  
  # calculate the pooled standard deviation
  pooled_sd <- sqrt(((n1 - 1) * s1 + (n2 - 1) * s2) / (n1 + n2 - 2))
  
  # calculate Cohen's d
  cohen_d <- (m1 - m2) / pooled_sd
  
  return(cohen_d)
}
```

### 3. Create a group variable in `mlm_df` called `morality_group`

We create a grouping variable called morality_group based on artists' morality scores.
We classify artists as "Bad" if their morality score is less than -1, "Good" if it's greater than 1, and NA for other cases.

```{r, echo = TRUE, eval = TRUE, comment = NA, message = FALSE, warning = FALSE}
# create grouping variable for `Morality`
art_morality <- art_morality %>%
  mutate(morality_group = case_when(
    Morality < -1 ~ "Bad",
    Morality > 1  ~ "Good",
    TRUE          ~ NA_character_ ## NA for other cases, or blank
  ))
```

### 4. Compute a *t*-test on `Beauty` using only the `"Bad"` and `"Good"` groups using *your* function

In this section, we perform a custom t-test to compare the Beauty ratings between the "Bad" and "Good" morality groups.

```{r}
#subset morality_group data into "bad" and "good" groups
bad_group <- art_morality %>%
  filter(morality_group == "Bad") %>%
  select(Beauty)
  
good_group <- art_morality %>%
  filter(morality_group == "Good") %>%
  select(Beauty)
```

```{r}
#perform t test using the function we created 
t_test(bad_group$Beauty, good_group$Beauty)
```

### 5. Compute a *t*-test on `Beauty` using only the `"Bad"` and `"Good"` groups using the `t.test` function (set `var.equal = TRUE`)

We also perform a t-test on the Beauty variable using the t.test function and compare the results with our custom t-test.

```{r, echo = TRUE, eval = TRUE, comment = NA, message = FALSE, warning = FALSE}
# t-test of equal variances with built-in R t.test function
t_test_result <- t.test(bad_group, good_group, var.equal = TRUE) # assumes that the variances of the two groups are equal, which is an assumption made in a standard two-sample t-test with equal variances

## printing result 
t_test_result
```

### 6. Compare your *t*-statistics from questions 4. and 5. Do they differ?

We compare the t-statistics from our custom t-test and the one performed using t.test to check for differences.

The p-values and degrees of freedom vary slightly from the custom t-test and the t.test.
Both come to the conclusion to reject the null hypothesis (based on the p-value being less than 0.05) and conclude that there is evidence to suggest a statistically significant difference in beauty ratings between the "Bad" and "Good" morality groups.

### 7. Intrepret the *p*-value from question 5. Were morally good paintings rated higher in beauty?

Since the p-value (0.01292) is less than the significance level (alpha = 0.05), we reject the null hypothesis.
This means there is evidence to suggest there is a statistically significant difference in beauty ratings between the "Bad" and "Good" morality groups.

Since our hypothesis is just checking for a difference in means, we cannot say that the good paintings are higher rated.
This can be guessed, or assumed, based on the sample estimate means, but that is not statistical evidence.

### 8. Compute Cohen's *d* using your function.

```{r, echo = TRUE, eval = TRUE, comment = NA, message = FALSE, warning = FALSE}
# Cohen's *d* using your function
cohens_d(bad_group$Beauty, good_group$Beauty)
```

The small magnitude of the Cohen's D value indicates that the effect of an artist's perceived morality on the beauty ratings of their artwork is minimal.
This means that, on average, participants' evaluations of the beauty of artworks are not substantially swayed by their views on the artist's morality.

### 9. Perform an ANOVA using `anova_test`

An ANOVA (Analysis of Variance) test is a statistical method used to compare the means of three or more groups to determine if there are any statistically significant differences between them.

```{r, echo = TRUE, eval = TRUE, comment = NA, message = FALSE, warning = FALSE}
anova_result <- anova_test(data = art_morality, dv = Beauty, between = morality_group)

# print results 
summary(anova_result)
```

The ANOVA results suggest that there is a statistically significant difference in beauty ratings between the "Bad" and "Good" morality groups.
The p-value is less than the typical significance level of 0.05, indicating that we can reject the null hypothesis.

### 10. Compute the square of the *t*-statistic from 5. How does it compare to the *F*-statistic in 9.?

```{r, echo = TRUE, eval = TRUE, comment = NA, message = FALSE, warning = FALSE}
# extract the t-statistic
t_statistic <- t_test_result$statistic

# compute the square of the t-statistic
t_stat_squared <- t_statistic

# print the result
t_stat_squared
```

The squared t-statistic is -2.487173 and the F-statistic is 6.186.
Let's interpret these values:

-   **Squared t-statistic:**

    -   The squared t-statistic measures the effect size of the difference in beauty ratings between the "Bad" and "Good" morality groups.

    -   A squared t-statistic of -2.487 indicates that the means of these two groups differ significantly

    -   The negative sign suggests that the "Bad" morality group has, on average, lower beauty ratings compared to the "Good" morality group

    -   The magnitude of the squared t-statistic (regardless of sign) indicates the strength of this difference.
        In our case, the magnitude suggests a moderate effect size.

-   **F-statistic:**

    -   The F-statistic from the ANOVA tests whether there is a statistically significant difference in mean beauty ratings between the "Bad" and "Good" morality groups.

    -   The F statistic quantifies the extent of this difference relative to the variability within the groups

    -   With an F-statistic of 6.186 and a p-value less than 0.05, it suggests that there is strong evidence to conclude that the mean beauty ratings differ significantly between these two groups.

In summary, both the squared t-statistic and the F-statistic provide evidence that the beauty ratings of paintings differ between the "Bad" and "Good" morality groups.

### 11. Perform an ANCOVA using `anova_test`

An Analysis of Covariance (ANCOVA) is used to compare means of a dependent variable across multiple groups while statistically controlling for the influence of one or more continuous covariates.

```{r warning=FALSE}
# perform ANCOVA using anova_test

ancova_result <- art_morality %>%
  anova_test(
    formula = Beauty ~ morality_group + AEQ + Attractiveness + Political + Skill, # specify the ANCOVA model formula
    dv = "Beauty", # dependent variable (DV) is "Beauty"
    between = "morality_group", # independent variable is "morality_group"
    covariate = c("AEQ", "Attractiveness", "Political", "Skill"), # specify covariates
    effect.size = "pes" # make the effect size to be computed as partial eta squared (n²)
  )

# print the ANCOVA summary
ancova_result
```

### 12. Interpret the ANCOVA in 11. Be sure to mention significant variables and their effect size (0.01 = small; 0.06 = moderate; 0.14 = large). Does including covariates change our interpretation?

*Let's interpret our results:*

-   Significant Predictor Variables:

    -   AEQ (aesthetic experiences) : The only significant predictor with a p-value of less than 0.05. This suggests that aesthetic experiences is a significant factor in how people rate the beauty of paintings.

-   Non-Significant Predictors:

    -   Morality Group, Attractiveness, Political Orientation, and Skill did not show significant effects (p \> 0.05).
        These factors do not seem to have a substantial effect on how people rate beauty in this context.

    -   The effect sizes for Attractiveness, Political orientation, and Skill are small, reinforcing their limited impact on beauty ratings.

-   Effect of Including Covariates:

    -   Including these covariates in the analysis does not substantially change the interpretation of how individuals rate the beauty of paintings.
        The small effect sizes, especially for non-significant variables, suggest that these factors have minimal influence.

    -   The significant effect of AEQ, however, indicates that it might have a unique contribution to how beauty is perceived in paintings, distinguishing it from the other variables tested.

In summary, the ANCOVA results suggest that among the variables tested, only AEQ (aesthetic experiences) has a significant impact on how individuals rate the beauty of paintings.
However, attributes like morality, physical attractiveness, political orientation, and skill level do not play a significant role in influencing beauty ratings.
The inclusion of covariates helps provide a more accurate understanding of the relationship between these predictors and Beauty ratings.

### 13. Does an artist's morality affect evaluations on their artwork?

To answer the research question, "Does an artist's morality affect evaluations of their artwork?" based on the analysis we've conducted, we can provide the following conclusion:

*Yes, an artist's morality does have an effect on the evaluations of their artwork, but the effect is not very strong.*

-   The t-tests and ANOVA conducted show that there is a statistically significant difference in the beauty ratings of artworks based on the morality group of the artist ("Bad" vs. "Good").
    This suggests that participants' evaluations of artwork are influenced to some extent by their perception of the artist's morality.

-   The Cohen's D value, which measures the effect size, indicates that the impact of the artist's morality on artwork evaluations is minimal.
    This means that while there is a statistically significant difference, it's not a strong or large effect.

-   Both ANOVA and ANCOVA results support the notion that there is a significant difference in beauty ratings between the morality groups.

-   The significant role of aesthethic experiences in the ANCOVA analysis indicates that aesthetic experiences might have a unique contribution to how beauty is perceived in paintings, which could potentially overshadow the impact of the artist's morality.

In summary, the analysis suggests that an artist's morality does affect evaluations of their artwork, but the effect is relatively modest and not as influential as other factors like the individual aesthetic experiences of the viewers.
